Hierarchical Controller Architecture in Software Defined Networks

### 2013-12-16 paper/paper.tex
references

The emergence of Software Defined Networks (SDN) has induced a separation between forwarding and control elements and has significantly improved the role of the control plane. The control elements may no longer be present on the switches but move to a separate device called a controller, and use predefined protocol such as OpenFlow [openflow-ref], to install forwarding rules. There may be a one-to-many mapping from controllers to switches. At one extreme, control plane may provide a control application with the global network view, which can then be implemented as a centralized application, rather than a distributed one. Regardless of this abstraction, the control plane itself must inevitably be a distributed system, with multiple controllers, to achieve desired level of scalability, reliability and responsiveness.

Conventional control plane architectures involve multiple peer controllers, where each controller operates on a subset of switches and maintains a global view of the entire network, which is exposed to the control application. We referred to such an architecture as \emph{flat controller architecture}. Underlying network state is shared with other controllers through distributed state dissemination frameworks such as distributed transactional databases or Distributed Hash Tables (DHT). Consequently, such an architecture results in huge communication overhead between controllers, which may hinders the scalability of the system. Besides, network state inconsistency is quite common in flat architecture which degrades the application performance significantly [logically ref TODO].

@@

\begin{figure}
\includegraphics[scale=0.5]{hierarchy}
\caption{Hierarchical arrangement of controllers}
\label{fig:hierarchy}
\end{figure}

@@

\subsection{Implementation}
\label{sec:implementation}
We wrote our own simulator, which is available at [TODO REF simulator]. We did not use an existing simulator/controller such as Mininet with Floodlight [TODO ref] for the following two reasons: (1) Existing controllers lack the support for flexible controller architecture and are not suitable for hierarchical arrangement of controllers. Modifying existing controllers such as Floodlight to add support for one-switch abstraction (section~\ref{subsec:aggr}), is not a good idea because it requires code/design changes in their primitive modules. Besides, we were not looking for a feature rich controller at that point of time. (2) Existing simulators do not provide controls to regulate replication between controllers or to measure communication overhead (number of messages exchanged ) between controllers, which we needed for evaluation of control plane architecture. Given the above requirements, writing a simulator which provides one-switch abstraction support for the controllers was the best option.

@@

To observe the effect of different architectures in real world applications, we used SWIM project [TODO ref swim], which provides synthetic representative test workloads  by sampling historical MapReduce cluster traces using [swim paper ref TODO]. The original trace comes from historical Hadoop traces on a 600 machine cluster at Facebook and spans $6$ months, and contains roughly $1$ million jobs. The representative trace file contains $\approx 6000$ jobs. The trace was simulated with a cluster of $600$ machines (as in the original cluster) connected through a network of $200$ switches and $10$ Gbps links. The cluster is divided into $10$ control domains, with $60$ hosts, $20$ switches and one controller for each domain. Each switch in a domain is connected to $1$ to $4$ switches within the same domain, with degree being chosen randomly. Each domain contains $2$ top-of-the-rack switches which inter-connect domains. For hierarchical architecture, there are $11$ controllers: $1$ controller per domain and $1$ root controller which manages the other $10$ controllers. While flat architecture consists of $10$ peer controllers, one for each domain.

@@

\subsection{Results}
\label{subsec:results}
We simulated the trace file for both flat and hierarchical architectures, with the same network topology and the same assignment of task nodes for each job. We calculated two metrics:
\begin{enumerate}
    \item The performance of control application which is measured as the length of the path (number of hops) assigned to flows (Figure ~\ref{fig:path}) and maximum link utilization encountered by the flow (Figure ~\ref{fig:util}). While the flat architecture gives optimal results, hierarchical architecture provides sub-optimal results.
    \item Communication overhead during flow set up, which is measured in term of number of messages passed between switches and controllers as well as within controllers (Figure ~\ref{fig:msg}). The simulation assumes instant replication of network state between controllers.
\end{enumerate}
Results show that hierarchical architecture reduces the communication overhead between controllers by an order of magnitude while modestly deviating from optimal solution.

@@

\subsection{Evaluation}
We used a dummy topology to show the benefits of timestamps. There are three controller domains 1, 2 and 3, with domain 2 connecting the other two domains. Flows arrive randomly between domain 1 and domain 3 and are routed through domain 2. Link Balancer Controller [logically - ref] runs on each controller as a control application: when a new flow arrives, a path which minimizes the maximum link utilization is assigned to the flow.

>>>>

The emergence of Software Defined Networks (SDN) has induced a separation between forwarding and control elements and has significantly improved the role of the control plane. The control elements may no longer be present on the switches but move to a separate device called a controller, and use predefined protocol such as OpenFlow \cite{openflow}, to install forwarding rules. There may be a one-to-many mapping from controllers to switches. At one extreme, control plane may provide a control application with the global network view, which can then be implemented as a centralized application, rather than a distributed one. Regardless of this abstraction, the control plane itself must inevitably be a distributed system, with multiple controllers, to achieve desired level of scalability, reliability and responsiveness.

Conventional control plane architectures involve multiple peer controllers, where each controller operates on a subset of switches and maintains a global view of the entire network, which is exposed to the control application. We referred to such an architecture as \emph{flat controller architecture}. Underlying network state is shared with other controllers through distributed state dissemination frameworks such as distributed transactional databases or Distributed Hash Tables (DHT). Consequently, such an architecture results in huge communication overhead between controllers, which may hinders the scalability of the system. Besides, network state inconsistency is quite common in flat architecture which degrades the application performance significantly \cite{logically-centralized}.

@@

\begin{figure}
\includegraphics[scale=0.5]{hierarchy}
\caption{A 2-layer Hierarchical arrangement of controllers: Root controller manages the layer of controllers, which in-turn manages a subset of switches.}
\label{fig:hierarchy}
\end{figure}

@@

\subsection{Implementation}
\label{sec:implementation}
We wrote our own simulator, which is available at \cite{simulator}. We did not use an existing simulator/controller such as Mininet\cite{mininet} with Floodlight\cite{floodlight} for the following two reasons: (1) Existing controllers lack the support for flexible controller architecture and are not suitable for hierarchical arrangement of controllers. Modifying existing controllers such as Floodlight to add support for one-switch abstraction (section~\ref{subsec:aggr}), is not a good idea because it requires code/design changes in their primitive modules. Besides, we were not looking for a feature rich controller at that point of time. (2) Existing simulators do not provide controls to regulate replication between controllers or to measure communication overhead (number of messages exchanged ) between controllers, which we needed for evaluation of control plane architecture. Given the above requirements, writing a simulator which provides one-switch abstraction support for the controllers was the best option.

@@

To observe the effect of different architectures in real world applications, we used SWIM project \cite{swim}, which provides synthetic representative test workloads  by sampling historical MapReduce cluster traces. The original trace comes from historical Hadoop traces on a 600 machine cluster at Facebook and spans $6$ months, and contains roughly $1$ million jobs. The representative trace file contains $\approx 6000$ jobs. The trace was simulated with a cluster of $600$ machines (as in the original cluster) connected through a network of $200$ switches and $10$ Gbps links. The cluster is divided into $10$ control domains, with $60$ hosts, $20$ switches and one controller for each domain. Each switch in a domain is connected to $1$ to $4$ switches within the same domain, with degree being chosen randomly. Each domain contains $2$ top-of-the-rack switches which inter-connect domains. For hierarchical architecture, there are $11$ controllers: $1$ controller per domain and $1$ root controller which manages the other $10$ controllers. While flat architecture consists of $10$ peer controllers, one for each domain.

@@

\subsection{Results}
\label{subsec:results}
We simulated the trace file for both flat and hierarchical architectures, with the same network topology and the same assignment of task nodes for each job. We calculated two metrics:
\begin{enumerate}
    \item The performance of control application which is measured as the length of the path (number of hops) assigned to flows (Figure ~\ref{fig:path}) and maximum link utilization encountered by the flow (Figure ~\ref{fig:util}). While the flat architecture gives optimal results, hierarchical architecture provides sub-optimal results. However, note that simulation provides instantaneous replication and thus eliminates network inconsistency between controller, which would  have otherwise, degraded the performance of flat architecture.
    \item Communication overhead during flow set up, which is measured in term of number of messages passed between switches and controllers as well as within controllers (Figure ~\ref{fig:msg}). The simulation assumes instant replication of network state between controllers.
\end{enumerate}
Results show that hierarchical architecture reduces the communication overhead between controllers by an order of magnitude while modestly deviating from optimal solution.

@@

\subsection{Evaluation}
We used a dummy topology to show the benefits of timestamps. There are three controller domains 1, 2 and 3, with domain 2 connecting the other two domains. Flows arrive randomly between domain 1 and domain 3 and are routed through domain 2. Link Balancer Controller \cite{logically-centralized} runs on each controller as a control application: when a new flow arrives, a path which minimizes the maximum link utilization is assigned to the flow.
